Open archaeology, open source?

### 2023-10-17 analysis/_03-data_methodology.qmd
Minor edits to the text.
Focused on the text concerning licensing. Fixed an innacurate encoding in Section 5. Updated the bib file. Removed irrelevant or completed TODOs and comments. Other very minor spelling and grammar tweaks.

We present an exploratory analysis of open-archaeo [[open-archaeo.info](open-archaeo.info), @batist2023], a directory of `r nrow(oarch)` pieces of open source archaeological software and other digital resources.

@@

We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, which are designed to facilitate open source software development by hosting public git repositories.
Git is a protocol designed to facilitate open and distributed contributions to a common code base. <!-- mention other version control systems like SVN and mercurial, maybe? -->
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a fork in which they apply their modifications, and then request that their fork be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent forks intact.

Git is also designed to facilitate code review.
All modifications are tracked as "diffs", which highlight additions or deletions to the code base, including changes within individual files.
Typically, a contributor will group a series of changes into a more comprehensive "commit" based on a specific task or part of a workflow.
Commits are always accompanied by a message, in which the contributor (ideally) describes the reason and context for the changes included in the commit.
Moreover, git assigns each commit a unique identifier and identifies the contributor by name and email address to ensure some degree of public accountability.

>>>>

We present an exploratory analysis of open-archaeo [[open-archaeo.info](open-archaeo.info), @batist2023], a directory of `r nrow(oarch)` <!-- TODO: make this consistent with what appears in Section 4, based on the filter --> pieces of open source archaeological software and other digital resources.

@@

We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, which are designed to facilitate open source software development by hosting public git repositories.
Git is a protocol designed to facilitate open and distributed contributions to a common code base.
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a fork in which they apply their modifications, and then request that their fork be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent forks intact.

Git is also designed to facilitate code review and version control.
All modifications are tracked as "diffs", which highlight additions or deletions to the code base, including changes within individual files.
Typically, a contributor will group a series of changes into a more comprehensive "commit" based on a specific task or part of a workflow.
Commits are always accompanied by a message, in which the contributor (ideally) describes the reason and context for the changes included in the commit.
Moreover, git assigns each commit a unique identifier and identifies the contributor by name and email address to ensure some degree of public accountability.

### 2023-10-17 analysis/_04-open_archaeology.qmd
Minor edits to the text.
Focused on the text concerning licensing. Fixed an innacurate encoding in Section 5. Updated the bib file. Removed irrelevant or completed TODOs and comments. Other very minor spelling and grammar tweaks.

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums.^[see https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code for a similar list of archaeology publications that include R code]
We annotated each record with categories and tags that describe what aspect of archaeological research each tool or resource was meant to address.
We categorized all records based on how each tool or resource is meant to be accessed or used, and assigned tags based on how developers identified their projects' purpose and scope.
See [batist2023] for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

To enact their mandate of ensuring that anyone can access and modify software and other creative works, the open source and open science movements encourage developers and scientists to adopt open licenses.
Licenses are legally-bindings statements that stipulate how a creative work can be accessed and used.
Proprietary licenses usually require explicit permission to be granted so that the work can be accessed or modified, usually in exchange for financial compensation.
Open licenses, on the other hand, are more permissive, and allow anyone to use creative works without such harsh restrictions.
While it is certainly possible to write your own license, it is very common to simply use one of several standardized open licenses (see [choosealicense.com](https://choosealicense.com for a more comprehensive overview of various options)).
Some licenses, like GNU, MIT and Apache, are explicitly suited for distributing software, and specify certain use cases that are afforded by digital media.
Other licenses, like the Creative Commons variants, are more suited to other kinds of creative works like books, articles, movies, music, photographs, and websites.
The Creative Commons licenses also include clauses that cater to academic or creative sensibilities, such as requirements to attribute credit to the original authors, to restrict commercial use, and to propagate similar restrictions in derivative works.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license [@tbl-repo-licenses].
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restricts they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of FOSS software within non-FOSS software, [@cite?]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@Balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@cite?].
Many repositories do not specify a license; their maintainers may have assumed that making the code available, without explicitly stating permitting use, is enough to allow unrestricted access to the repository's contents.
This is evocative of a similar ignorance among academics regarding the unsustainable use of GitHub as a long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023].
<!-- TODO: Seems like a pair of slightly uncharitable assumptions? -JR -->
<!-- TODO: (ZB) Expand with additional info after seeing what is in the "other" category -->

>>>>

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
It includes both software and documents, but not research compendiums.^[see [https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code](https://github.com/benmarwick/ctv-archaeology#publications-that-include-r-code) for a similar list of archaeology publications that include R code]
We annotated each record with categories and tags that describe what aspect of archaeological research each tool or resource was meant to address.
We categorized all records based on how each tool or resource is meant to be accessed or used, and assigned tags based on how developers identified their projects' purpose and scope.
See [@batist2023] for a more comprehensive overview of the tags and categories applied to open-archaeo.

@@

To enact their mandate of ensuring that anyone can access and modify software and other creative works, the open source and open science movements encourage developers and scientists to adopt open licenses.
Licenses are legally-binding statements that stipulate how a creative work can be accessed and used.
Proprietary licenses usually require explicit permission to be granted so that the work can be accessed or modified, usually in exchange for financial compensation.
Open licenses, on the other hand, are more permissive, and allow anyone to use creative works without such harsh restrictions.
While it is certainly possible to write your own license, it is very common to simply use one of several standardized open licenses (see [choosealicense.com](https://choosealicense.com for a more comprehensive overview of various options)).
Some licenses, like GNU, MIT and Apache, are explicitly suited for distributing software, and specify certain use cases that are afforded by digital media.
Other licenses, like the Creative Commons variants, are more suited to other kinds of creative works like books, articles, movies, music, photographs, and websites.
The Creative Commons licenses also include clauses that cater to academic or creative sensibilities, such as requirements to attribute credit to the original authors, to restrict commercial use, and to propagate similar restrictions in derivative works.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license [@tbl-repo-licenses].
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of FOSS software within non-FOSS software, [@dusollier2007]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

### 2023-10-17 analysis/_05-collaboration.qmd
Minor edits to the text.
Focused on the text concerning licensing. Fixed an innacurate encoding in Section 5. Updated the bib file. Removed irrelevant or completed TODOs and comments. Other very minor spelling and grammar tweaks.

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_50)`.
This may be attributed to the steep learning curve commonly attributed to working with git.
While git can be a great way to track changes and manage distributed contributions to a common code base, it can also be unwieldy in situations when multiple users (especially those with less experience using git for collaborative purposes) are expected to contribute within short spans of time.
This meshes with our prior observations that projects which tend to exhibiting higher commit rates have less contributors.
Additionally, our analyses neglects to account for contributions that are not tracked via git or GitHub.
Those who do not code may provide creative guidance or feedback during in-person meetings, via email, or using alternative online messaging or social media platforms.
A more focused qualitative assessment of these non-coding and supportive work practices would shed more light on the totality of effort that goes into producing and maintaining open source projects.

<!-- TODO: Abrupt transition, need to revise -->
The prototypical and idyllic collaborative open source project ingrained into the common consciousness hypothetically comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects (<!-- TODO: do these share any characteristics in terms of platform or subject? -->) operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.

>>>>

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_75)`.
This may be attributed to the steep learning curve commonly attributed to working with git.
While git can be a great way to track changes and manage distributed contributions to a common code base, it can also be unwieldy in situations when multiple users (especially those with less experience using git for collaborative purposes) are expected to contribute within short spans of time.
This meshes with our prior observations that projects which tend to exhibiting higher commit rates have less contributors.
Additionally, our analyses neglects to account for contributions that are not tracked via git or GitHub.
Those who do not code may provide creative guidance or feedback during in-person meetings, via email, or using alternative online messaging or social media platforms.
A more focused qualitative assessment of these non-coding and supportive work practices would shed more light on the totality of effort that goes into producing and maintaining open source projects.

The prototypical and idyllic collaborative open source project ingrained into the common consciousness hypothetically comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.
